Neural Sparse Gaussian Process: For Updated Exploration

### 2023-04-14 paper/main.tex
Switch to use bibtex and update Makefile

% NeurIPS packages
\usepackage[preprint,nonatbib]{neurips_2022}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
%\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors

@@

@@

\cite{hernandez2016scalable}
% \section{dual SVGP}
% In this section we extend these fast updates to environment's with high dimensional state spaces and large data sets.
% % Our method draws on the connection between BNNs and GPs and formulates a function space SVGP posterior given
% Our method uses a BNN dynamic model and draws on the connection between BNNs and GPs
% \parencite{khanApproximate2019} to formulate a function space SVGP posterior,
% where we can apply the fast updates from \cref{eq-dual-update-svgp}.
% At a high-level, we first use Laplace's approximation to obtain a weight space posterior for our BNN.
% We then linearise our BNN around the optimal parameters and interpret it as a GP.
% Finally, we formulate a lower rank approximation of this GP posterior (i.e. a SVGP posterior) using inducing variables.
% % formulates a function space SVGP posterior by drawing on the connection between BNNs and GPs.

@@

\textbf{Sparse Gaussian processes}
% The GP predictive posterior in \cref{eq-gp-predictive-posterior} is computationally expensive as it requires inverting an $N\times N$ matrix.
A practical and computationally appealing alternative is to augment the joint probability space with pseudo inputs $\mathbf{Z} \in \R^{M \times D}$
and corresponding outputs
$\mathbf{u} = f(\mathbf{Z}; \transitionParams) \in \R^{M \times C}$, known as inducing variables, where $M \ll N$.
\todo{cite sparse gp papers}
Instead of collapsing the inducing variables like \cite{titsiasVariational2009}, they can be treated variationally
$p(\mathbf{u} \mid \mathbf{y}) \approx q(\mathbf{u}) = \mathcal{N}\left( \mathbf{u} \mid \mathbf{m}, \mathbf{V} \right)$, where $\mathbf{m}$ and $\mathbf{V}$ are variational parameters
which need to be optimised \parencite{hensmanGaussian2013}.
The sparse variational GP (SVGP) predictive posterior is then given by,
\begin{align}
\label{eq-dual-svgp-predictive-posterior}
  p(f(\mathbf{x}_{*}) \mid \mathbf{y})
&\approx \mathcal{N} \left( f(\mathbf{x}_{*}) \mid \mathbf{k}_{*\mathbf{z}} \mathbf{K}^{-1}_{\mathbf{z}\mathbf{z}} \mathbf{m}^{*},
  k_{**} - \mathbf{k}_{*\mathbf{z}} \mathbf{K}^{-1}_{\mathbf{z}\mathbf{z}} \mathbf{k}_{*\mathbf{z}}^{T}
  % \mathbf{k}_{*\inducingInput} \mathbf{K}^{-1}_{\mathbf{z}\mathbf{z}} \mathbf{k}_{*\inducingInput}^{T}
  + \mathbf{k}_{*\mathbf{z}} \mathbf{K}^{-1}_{\mathbf{z}\mathbf{z}}  \mathbf{V}^{*}  \mathbf{K}^{-1}_{\mathbf{z}\mathbf{z}} \mathbf{k}_{*\mathbf{z}}^{T}
  \right)
\coloneqq q_{\inducingVariable}(f(\mathbf{x}_{*}))
\end{align}

@@

\textbf{NN to GP}
Given this interpretation, we can represent the weight space prior from \cref{eq-log-prior}, in
function space, by linearising the BNN around $\transitionParams^{*}$ and interpretting it as a GP.
The GP prior is then given by,
\begin{align} \label{eq-laplace-approx-function-space}
  % \transitionFn_{\transitionParams_{i}}(\cdot) \sim \mathcal{N} \left( \mu_{\transitionParams_{i}^{*}}(\cdot), K_{\transitionParams_{i}^{*}}(\cdot, \cdot') \right) \qquad
  f(\cdot ;\transitionParams) \sim \mathcal{N} \left( \mu(\cdot), k(\cdot, \cdot') \right) \quad
  % \mu_{\transitionParams_{i}^{*}}(\cdot)
  \mu(\cdot)
  = 0 \quad
  % = f(\cdot ;\transitionParams^{*}) \quad
  % K_{\transitionParams_{i}^{*}}(\cdot, \cdot')
  k(\cdot, \cdot')
  = \frac{1}{\gamma^{2}} \mathbf{J}(\cdot) \mathbf{J}(\cdot')^{T},
\end{align}
where the kernel is the neural tangent kernel (NTK) \parencite{immerImprovingPredictionsBayesian2021},
\todo{what mean function shoudl this use???}
\todo{NTK times some factor??}
with Jacobian given by $\mathbf{J}(\cdot) = \odv{\transitionFn(\cdot; \transitionParams)}{\transitionParams}_{\transitionParams=\transitionParams^{*}}$.
% The kernel in \cref{eq-laplace-approx-function-space} is the neural tangent kernel (NTK)
% $k(\mathbf{x}, \mathbf{x}') = \sigma_{0}^{2} J_{\transitionParams^{*}}(\mathbf{x}) J_{\transitionParams^{*}}(\mathbf{x}')^{T}$
% \parencite{immerImprovingPredictionsBayesian2021}
% to formulate the GP posterior $p(f(\mathbf{X} ;\transitionParams_{0:i}) \mid \dataset_{0:i})$.
Given this GP prior, we can use \cref{eq-gp-predictive-posterior} to obtain our GP posterior by conditioning on the data.
% Given our GP prior, we can use the properties of multivariate normals to obtain our GP posterior by conditioning on the data.
% where $\mathbf{A} = \mathbf{k}_{*\mathbf{x}} \mathbf{K}^{-1}_{\mathbf{x}\mathbf{x}}$
In contrast to conventional GP kernels, the NTK does not have any hyperparameters which need to be learned.
Intuitively, we can view the optimisation in \cref{eq-empirical-risk} as learning our kernel.
As a result, our NTK may be highly non stationary as it is dependent on the NN architecture, for example, the activation functions.
It is worth noting that the NTK linearises the network around the optimised parameters $\transitionParams^{*}$,
so the function space prior (and thus posterior) is only a locally linear approximation.

@@

We follow \cite{csatoSparseOnlineGaussian2002} and express our GP posterior in the dual parameter space.
However, we consider the variational formulation introduced by \cite{adamDualParameterizationSparse2021}, where the optimal variational parameters are given by,
\begin{align} \label{eq-dual-params}
\mathbf{m}^{*} = \mathbf{V}^{*}\bm\alpha^{*} \quad \mathbf{V}^{*} = [\mathbf{K}_{\mathbf{z}\mathbf{z}}^{-1} + \bm\beta^{*}]
\end{align}
Importantly, we can calculate the natural parameters $(\bm\alpha, \bm\beta)$ in closed form,
\todo{paul needs to update \cref{eq-dual-svgp-params}}
\begin{align} \label{eq-dual-svgp-params}
\bm\alpha_{n} &=  \nabla_{\meanParam{1}} \mathbb{E}_{q_{\inducingVariable}(f(\mathbf{x}_{n}))} \left[ \log p(\mathbf{y}_{n} \mid f(\mathbf{x}_{n}) ) \right] \\
\bm\beta_{n} &=  \nabla_{\meanParam{1}} \mathbb{E}_{q_{\inducingVariable}(f(\mathbf{x}_{n}))} \left[ \log p(\mathbf{y}_{n} \mid f(\mathbf{x}_{n}) ) \right].
\end{align}
This simplifies SVGP inference \parencite{hensmanGaussian2013} as it no longer requires the variational parameters $(\mathbf{m}, \mathbf{V})$ to be optimised.

@@

\textbf{Hallucinated upper confidence RL}
A more theoretically grounded exploration strategy is UCRL \autocite{jakschNearoptimal2010}, which optimises joinly over
policies and models inside the set
\(\mathcal{M} = \{ f \mid | f(s,a) - \mu_{i}(s, a) | \leq \beta_{i} \Sigma_{i}(s, a) \quad \forall s, a \in \mathcal{S} \times \mathcal{A} \}\), representing all statistically plausible
models under the posterior \(p(f(s,a) \mid \mathcal{D}_{0:i} \cup (s,a)) = \mathcal{N}(f(s,a) \mid \mu_{i}(s,a), \Sigma_{i}(s,a))\) at episode \(i\).
This strategy is given by,
\begin{align} \label{eq-ucrl}
\policy_{i+1}^{\text{UCRL}} = \arg \max_{\policy \in \policyDomain} \max_{\transitionFn \in \mathcal{M}} J(\transitionFn, \policy).
\end{align}
This strategy optimises an optimistic policy over the set of plausible dynamics models.
Although this joint optimisation is intractable in general,
\cite{curiEfficient2020} proposed a practical alternative which is detailed in \cref{sec-hucrl}.

@@

In this section we extend these fast updates to environment's with high dimensional state spaces and large data sets.
% Our method draws on the connection between BNNs and GPs and formulates a function space SVGP posterior given
Our method uses a BNN dynamic model and draws on the connection between BNNs and GPs
\parencite{khanApproximate2019} to formulate a function space SVGP posterior,
where we can apply the fast updates from \cref{eq-dual-update-svgp}.
At a high-level, we first use Laplace's approximation to obtain a weight space posterior for our BNN.
We then linearise our BNN around the optimal parameters and interpret it as a GP.
Finally, we formulate a lower rank approximation of this GP posterior (i.e. a SVGP posterior) using inducing variables.
% formulates a function space SVGP posterior by drawing on the connection between BNNs and GPs.

@@

\subsection{Biblatex}
Rember when using biblatex to use 'parencite' for \parencite{kamtheDataEfficient2018} and when using natbib to use 'citep'.

>>>>

% Pass options to natbib
\PassOptionsToPackage{numbers, compress}{natbib}

% NeurIPS packages
\usepackage[preprint]{neurips_2022}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
%\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors

@@

% Let's use good old bibtex instead

@@

\cite{hernandez2016scalable}
% \section{dual SVGP}
% In this section we extend these fast updates to environment's with high dimensional state spaces and large data sets.
% % Our method draws on the connection between BNNs and GPs and formulates a function space SVGP posterior given
% Our method uses a BNN dynamic model and draws on the connection between BNNs and GPs
% \citep{khanApproximate2019} to formulate a function space SVGP posterior,
% where we can apply the fast updates from \cref{eq-dual-update-svgp}.
% At a high-level, we first use Laplace's approximation to obtain a weight space posterior for our BNN.
% We then linearise our BNN around the optimal parameters and interpret it as a GP.
% Finally, we formulate a lower rank approximation of this GP posterior (i.e. a SVGP posterior) using inducing variables.
% % formulates a function space SVGP posterior by drawing on the connection between BNNs and GPs.

@@

\textbf{Sparse Gaussian processes}
% The GP predictive posterior in \cref{eq-gp-predictive-posterior} is computationally expensive as it requires inverting an $N\times N$ matrix.
A practical and computationally appealing alternative is to augment the joint probability space with pseudo inputs $\mathbf{Z} \in \R^{M \times D}$
and corresponding outputs
$\mathbf{u} = f(\mathbf{Z}; \transitionParams) \in \R^{M \times C}$, known as inducing variables, where $M \ll N$.
\todo{cite sparse gp papers}
Instead of collapsing the inducing variables like \cite{titsiasVariational2009}, they can be treated variationally
$p(\mathbf{u} \mid \mathbf{y}) \approx q(\mathbf{u}) = \mathcal{N}\left( \mathbf{u} \mid \mathbf{m}, \mathbf{V} \right)$, where $\mathbf{m}$ and $\mathbf{V}$ are variational parameters
which need to be optimised \citep{hensmanGaussian2013}.
The sparse variational GP (SVGP) predictive posterior is then given by,
\begin{align}
\label{eq-dual-svgp-predictive-posterior}
  p(f(\mathbf{x}_{*}) \mid \mathbf{y})
&\approx \mathcal{N} \left( f(\mathbf{x}_{*}) \mid \mathbf{k}_{*\mathbf{z}} \mathbf{K}^{-1}_{\mathbf{z}\mathbf{z}} \mathbf{m}^{*},
  k_{**} - \mathbf{k}_{*\mathbf{z}} \mathbf{K}^{-1}_{\mathbf{z}\mathbf{z}} \mathbf{k}_{*\mathbf{z}}^{T}
  % \mathbf{k}_{*\inducingInput} \mathbf{K}^{-1}_{\mathbf{z}\mathbf{z}} \mathbf{k}_{*\inducingInput}^{T}
  + \mathbf{k}_{*\mathbf{z}} \mathbf{K}^{-1}_{\mathbf{z}\mathbf{z}}  \mathbf{V}^{*}  \mathbf{K}^{-1}_{\mathbf{z}\mathbf{z}} \mathbf{k}_{*\mathbf{z}}^{T}
  \right)
\coloneqq q_{\inducingVariable}(f(\mathbf{x}_{*}))
\end{align}

@@

\textbf{NN to GP}
Given this interpretation, we can represent the weight space prior from \cref{eq-log-prior}, in
function space, by linearising the BNN around $\transitionParams^{*}$ and interpretting it as a GP.
The GP prior is then given by,
\begin{align} \label{eq-laplace-approx-function-space}
  % \transitionFn_{\transitionParams_{i}}(\cdot) \sim \mathcal{N} \left( \mu_{\transitionParams_{i}^{*}}(\cdot), K_{\transitionParams_{i}^{*}}(\cdot, \cdot') \right) \qquad
  f(\cdot ;\transitionParams) \sim \mathcal{N} \left( \mu(\cdot), k(\cdot, \cdot') \right) \quad
  % \mu_{\transitionParams_{i}^{*}}(\cdot)
  \mu(\cdot)
  = 0 \quad
  % = f(\cdot ;\transitionParams^{*}) \quad
  % K_{\transitionParams_{i}^{*}}(\cdot, \cdot')
  k(\cdot, \cdot')
  = \frac{1}{\gamma^{2}} \mathbf{J}(\cdot) \mathbf{J}(\cdot')^{T},
\end{align}
where the kernel is the neural tangent kernel (NTK) \citep{immerImprovingPredictionsBayesian2021},
\todo{what mean function shoudl this use???}
\todo{NTK times some factor??}
with Jacobian given by $\mathbf{J}(\cdot) = \odv{\transitionFn(\cdot; \transitionParams)}{\transitionParams}_{\transitionParams=\transitionParams^{*}}$.
% The kernel in \cref{eq-laplace-approx-function-space} is the neural tangent kernel (NTK)
% $k(\mathbf{x}, \mathbf{x}') = \sigma_{0}^{2} J_{\transitionParams^{*}}(\mathbf{x}) J_{\transitionParams^{*}}(\mathbf{x}')^{T}$
% \citep{immerImprovingPredictionsBayesian2021}
% to formulate the GP posterior $p(f(\mathbf{X} ;\transitionParams_{0:i}) \mid \dataset_{0:i})$.
Given this GP prior, we can use \cref{eq-gp-predictive-posterior} to obtain our GP posterior by conditioning on the data.
% Given our GP prior, we can use the properties of multivariate normals to obtain our GP posterior by conditioning on the data.
% where $\mathbf{A} = \mathbf{k}_{*\mathbf{x}} \mathbf{K}^{-1}_{\mathbf{x}\mathbf{x}}$
In contrast to conventional GP kernels, the NTK does not have any hyperparameters which need to be learned.
Intuitively, we can view the optimisation in \cref{eq-empirical-risk} as learning our kernel.
As a result, our NTK may be highly non stationary as it is dependent on the NN architecture, for example, the activation functions.
It is worth noting that the NTK linearises the network around the optimised parameters $\transitionParams^{*}$,
so the function space prior (and thus posterior) is only a locally linear approximation.

@@

We follow \cite{csatoSparseOnlineGaussian2002} and express our GP posterior in the dual parameter space.
However, we consider the variational formulation introduced by \cite{adamDualParameterizationSparse2021}, where the optimal variational parameters are given by,
\begin{align} \label{eq-dual-params}
\mathbf{m}^{*} = \mathbf{V}^{*}\bm\alpha^{*} \quad \mathbf{V}^{*} = [\mathbf{K}_{\mathbf{z}\mathbf{z}}^{-1} + \bm\beta^{*}]
\end{align}
Importantly, we can calculate the natural parameters $(\bm\alpha, \bm\beta)$ in closed form,
\todo{paul needs to update \cref{eq-dual-svgp-params}}
\begin{align} \label{eq-dual-svgp-params}
\bm\alpha_{n} &=  \nabla_{\meanParam{1}} \mathbb{E}_{q_{\inducingVariable}(f(\mathbf{x}_{n}))} \left[ \log p(\mathbf{y}_{n} \mid f(\mathbf{x}_{n}) ) \right] \\
\bm\beta_{n} &=  \nabla_{\meanParam{1}} \mathbb{E}_{q_{\inducingVariable}(f(\mathbf{x}_{n}))} \left[ \log p(\mathbf{y}_{n} \mid f(\mathbf{x}_{n}) ) \right].
\end{align}
This simplifies SVGP inference \citep{hensmanGaussian2013} as it no longer requires the variational parameters $(\mathbf{m}, \mathbf{V})$ to be optimised.

@@

\textbf{Hallucinated upper confidence RL}
A more theoretically grounded exploration strategy is UCRL \citep{jakschNearoptimal2010}, which optimises joinly over
policies and models inside the set
\(\mathcal{M} = \{ f \mid | f(s,a) - \mu_{i}(s, a) | \leq \beta_{i} \Sigma_{i}(s, a) \quad \forall s, a \in \mathcal{S} \times \mathcal{A} \}\), representing all statistically plausible
models under the posterior \(p(f(s,a) \mid \mathcal{D}_{0:i} \cup (s,a)) = \mathcal{N}(f(s,a) \mid \mu_{i}(s,a), \Sigma_{i}(s,a))\) at episode \(i\).
This strategy is given by,
\begin{align} \label{eq-ucrl}
\policy_{i+1}^{\text{UCRL}} = \arg \max_{\policy \in \policyDomain} \max_{\transitionFn \in \mathcal{M}} J(\transitionFn, \policy).
\end{align}
This strategy optimises an optimistic policy over the set of plausible dynamics models.
Although this joint optimisation is intractable in general,
\cite{curiEfficient2020} proposed a practical alternative which is detailed in \cref{sec-hucrl}.

@@

In this section we extend these fast updates to environment's with high dimensional state spaces and large data sets.
% Our method draws on the connection between BNNs and GPs and formulates a function space SVGP posterior given
Our method uses a BNN dynamic model and draws on the connection between BNNs and GPs
\citep{khanApproximate2019} to formulate a function space SVGP posterior,
where we can apply the fast updates from \cref{eq-dual-update-svgp}.
At a high-level, we first use Laplace's approximation to obtain a weight space posterior for our BNN.
We then linearise our BNN around the optimal parameters and interpret it as a GP.
Finally, we formulate a lower rank approximation of this GP posterior (i.e. a SVGP posterior) using inducing variables.
% formulates a function space SVGP posterior by drawing on the connection between BNNs and GPs.

@@

\subsection{Biblatex}
Rember when using biblatex to use 'parencite' for \citep{kamtheDataEfficient2018} and when using natbib to use 'citep'.
